---
title: "Mental_Illness.Mar20_Feb21.Rmd"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(haven)
library(ggplot2)
library(pheatmap)
library(fastDummies)
library(glmnet)
library(coefplot)
```

## Project Goals

Replicate Mental Health By Political Affiliation Study by Zach Goldberg (https://threadreaderapp.com/thread/1248823584111439872.html)

Zach Goldberg demonstrated that liberals are much more likely to report having been
diagnosed with a mental illness in a pew research survey from March 2020 (https://www.pewresearch.org/social-trends/dataset/covid-19-late-march-2020/)

A stated limitation of Goldberg's analysis was: 'It's possible that the disparities in self-reported diagnosis are simply or partly a function of white liberals being more likely to seek mental health evaluations. I don't have the data to answer this question.'

However, the data to answer that question was available in the Pew questionnaire.  In addition to asking, 'Has a doctor or other healthcare provider EVER told you that you have a mental health condition ?' the survey also asked:

1. In the past 7 days, how often have you…   Felt nervous, anxious, or on edge?
2. In the past 7 days, how often have you…   Felt depressed?
3. In the past 7 days, how often have you…   Felt lonely?
4. In the past 7 days, how often have you…   Felt hopeful about the future?
5. In the past 7 days, how often have you…   Had trouble sleeping?

Each of these was scored 1-5 with higher scores indicating more frequent distress (D. was reversed) and aggregated in a variable titled 'MHEALTH'.

These questions were asked again one year later in the W83 survey (https://www.pewresearch.org/science/dataset/american-trends-panel-wave-83/), providing opportunity the replicate the findings

### Specific Goals:
1. Independently replicate Goldberg's findings.
2. Repeat analysis using self-reported mental health variables (MHEALTH).
3. Replicate findings in another data set (W83).


## Step 1: Load the original W64 data set
```{r LoadData.W64}
mar20Dat <- haven::read_sav(file=file.path('W64_Mar20/', 'ATP W64.sav'))

#Make a working copy
mar20DF <- as.data.frame(mar20Dat)
```

## Step 2: Re-encode variables as necessary

Main Variables
"F_IDEO": Ideology
Labels:
 value             label
     1 Very conservative
     2      Conservative
     3          Moderate
     4           Liberal
     5      Very liberal
     
"COVID_MENTAL_W64":  "Has a doctor or other healthcare provider EVER told you that you have a mental health condition ?"   
"MHEALTH": "Simple additive index of 5 mental health measures"

Relevant covariates
"F_ACSWEB": "ACS Internet Usage"
"F_AGECAT": "Age category"
"F_MARITAL": "Marital status"
"F_EDUCCAT2": "Education level category 2" 
"F_INCOME": "Family income"
"F_ATTEND": "Religious service attendance"
"F_RACETHN": "Race-Ethnicity"
"F_SEX": "Sex"

Re-encoded so that a higher score means more of the thing measured (for example, a higher MARITAL score means married) and variables such as RACE become encoded as YES/NO answers to individual races (For example: Hispanic).

```{r LoadRelevantVars}
#Note: Race will need to be one-hot
relCov <- c("F_ACSWEB", "F_AGECAT", "F_MARITAL", "F_EDUCCAT2", "F_INCOME",  "F_IDEO", "F_ATTEND")  #"MHEALTH", "F_RACETHN" , "F_SEX", "COVID_MENTAL_W64"

mar20Dat.curDF <- as.data.frame(mar20Dat[, c(relCov, "MHEALTH", "F_RACETHN", "F_SEX", "COVID_MENTAL_W64")])
mar20Dat.curDF[mar20Dat.curDF==99] <- as.numeric(NA)
mar20Dat.curDF <- mar20Dat.curDF[!apply(mar20Dat.curDF, 1, function(x){any(is.na(x))}),]

#One-hot race
F_RACETHN_Name <- names(attr(mar20Dat$F_RACETHN, "labels"))[match(mar20Dat.curDF$F_RACETHN, attr(mar20Dat$F_RACETHN, "labels"))]
table(F_RACETHN_Name)
oHot_Race <- fastDummies::dummy_cols(F_RACETHN_Name, remove_selected_columns=TRUE)
colnames(oHot_Race) <- make.names(sub('\\.data_', '', colnames(oHot_Race)))

#One-hot sex
F_SEX_Name <- names(attr(mar20Dat$F_SEX, "labels"))[match(mar20Dat.curDF$F_SEX, attr(mar20Dat$F_SEX, "labels"))]
table(F_SEX_Name)
oHot_Sex <- fastDummies::dummy_cols(F_SEX_Name, remove_selected_columns=TRUE)
colnames(oHot_Sex) <- make.names(sub('\\.data_', '', colnames(oHot_Sex)))

oHots <- cbind(oHot_Race, oHot_Sex)

#Fix religion and marriage so higher is better
mar20Dat.curDF <- cbind(mar20Dat.curDF, oHots)
mar20Dat.curDF$F_HOW_MUCH_RELIGION <- 7-mar20Dat.curDF$F_ATTEND

mar20Dat.curDF$F_HOW_MUCH_MARRIED <- 7-mar20Dat.curDF$F_MARITAL

#Fix mental illness diagnosis so 1 is TRUE and 0 is FALSE
mar20Dat.curDF$DIAG_MENTAL_ILLNESS <- 2-mar20Dat.curDF$COVID_MENTAL_W64

#Fix MHEALTH so it varies by percentage of mental distress
mar20Dat.curDF$MENTAL_DISTRESS_PERCENT <- round(100*(mar20Dat.curDF$MHEALTH - min(mar20Dat.curDF$MHEALTH))/(max(mar20Dat.curDF$MHEALTH)- min(mar20Dat.curDF$MHEALTH)))
```

## Step 3: Replicate Diagnosed Mental Illness Results

If Zach's observations hold up, the more liberal a person is, the more likely they are to have been diagnosed with a mental illness.
```{r DiagMentalIll.table}
diagMentalTable <- table(mar20Dat.curDF$DIAG_MENTAL_ILLNESS, mar20Dat.curDF$F_IDEO)
rownames(diagMentalTable) <- c('NOT.DIAGNOSED', 'DIAG.MENTAL.ILL')
colnames(diagMentalTable) <- names(attr(mar20Dat$F_IDEO, "labels"))[match(colnames(diagMentalTable), attr(mar20Dat$F_IDEO, "labels"))]
#knitr::kable(diagMentalTable)
diagMentalTable

diagMentalTable.per <- apply(diagMentalTable, 2, function(x){round(100*x/sum(x),1)})
#knitr::kable(diagMentalTable.per)
diagMentalTable.per

diagTable.per <- diagTable <- list()
for(curAge in unique(mar20Dat.curDF$F_AGECAT)) {
  ageName <- names(attr(mar20Dat.curDF$F_AGECAT, 'labels'))[attr(mar20Dat.curDF$F_AGECAT, 'labels') == curAge]
  ageName <- paste(curAge, ageName, sep='_')
  curIdx <- mar20Dat.curDF$F_AGECAT == curAge

  dMT <- table(mar20Dat.curDF[curIdx,'DIAG_MENTAL_ILLNESS'], mar20Dat.curDF[curIdx,'F_IDEO'])
  rownames(dMT) <- c('NOT.DIAGNOSED', 'DIAG.MENTAL.ILL')
  colnames(dMT) <- names(attr(mar20Dat$F_IDEO, "labels"))[match(colnames(dMT), attr(mar20Dat$F_IDEO, "labels"))]

  dMT.per <- apply(dMT, 2, function(x){round(100*x/sum(x),1)})
  
  diagTable[[ageName]] <- dMT
  diagTable.per[[ageName]] <- dMT.per
}
diagTable.per <- diagTable.per[order(names(diagTable.per))]
message('Percentage with diagnosed mentall illness by Age catagory')
diagTable.per
```

```{r DiagMentalIll.plot}
mar20Dat.curDF$F_IDEO_Name <- names(attr(mar20Dat$F_IDEO, "labels"))[match(mar20Dat.curDF$F_IDEO, attr(mar20Dat$F_IDEO, "labels"))]
means <- tapply(mar20Dat.curDF$DIAG_MENTAL_ILLNESS, mar20Dat.curDF$F_IDEO_Name, mean, na.rm=TRUE)
means <- sort(means)

plot(means*100, main='Mental Illness by Ideology', xaxt = "n", xlab = 'Ideology', ylab = "% Diagnosed Mentally Ill", type = 'o')
axis(1, at = 1:length(means), labels = names(means))
```

## Step 4: Repeat analysis using mental health score.

MHEALTH is an measure of mental distress over the last 7 days. It varies between 5 (least distressed) and 20 (most distressed).  This is converted to a percentage of the maximum amount of mental distress

```{r PlotMHscoress}
message('Distribution of mental distress scores')
summary(mar20Dat.curDF$MHEALTH)

message('Distribution of mental distress percentages of max')
summary(mar20Dat.curDF$MENTAL_DISTRESS_PERCENT)

mar20Dat.curDF$F_IDEO_Name <- names(attr(mar20Dat$F_IDEO, "labels"))[match(mar20Dat.curDF$F_IDEO, attr(mar20Dat$F_IDEO, "labels"))]
means <- tapply(mar20Dat.curDF$MENTAL_DISTRESS_PERCENT, mar20Dat.curDF$F_IDEO_Name, mean, na.rm=TRUE)
mar20Dat.curDF$F_IDEO_Name <- factor(mar20Dat.curDF$F_IDEO_Name,levels=names(means)[order(means)])

p <- ggplot(mar20Dat.curDF, aes(x = F_IDEO_Name, y = MENTAL_DISTRESS_PERCENT))
p <- p + geom_boxplot()
p <- p + stat_summary(fun=mean, shape=1,col='red', geom='point')
print(p)
```
Unlike the diagnosed mental illness (which seems to be following a hockey stick shape), the mental distress score seems to follow a linear relationship.  However, the same pattern holds: the more liberal you are, the more symptoms of mental illness you report. 

## Step 5: Control for covariates

One might expect that someone who has a very low income might naturally suffer from greater mental distress due to poverty as well as identify with a more liberal ideology since liberals tend to want to provide more services to low income people.  Therefore, it will be necessary to control for different covariates such as income.

```{r Covariates.regression}
eq1 <- sub('F_MARITAL', 'F_HOW_MUCH_MARRIED', sub('F_ATTEND', 'F_HOW_MUCH_RELIGION', paste(relCov, collapse=' + ')))
eq2 <- paste(colnames(oHots)[!colnames(oHots) %in% c('Other', 'Male')], collapse=' + ')
fullEq.str <- paste0('MENTAL_DISTRESS_PERCENT ~ ', eq1, " + ", eq2, " + DIAG_MENTAL_ILLNESS")

message('Prediction of MENTAL_DISTRESS_PERCENT aggregate variable')
message(fullEq.str)
fullEq <- as.formula(fullEq.str)

fullReg <- glm(formula = fullEq, data = mar20Dat.curDF)
summary(fullReg)

regPred <- predict(fullReg)
r2 <- cor(regPred, mar20Dat.curDF$MHEALTH)^2
message(paste('R2 =', round(r2,3)))
```

If you're not used to looking at results like these they can be intimidating.  Here's the important parts: 

Coefficients:
                    Estimate Std. Error t value Pr(>|t|)    
F_AGECAT            -2.44557    0.21189 -11.542  < 2e-16 ***

The first column is the variable name.  The column labeled 'Pr(>|t|)' is a p-Value indicating how significant that variable is for predicting MENTAL_DISTRESS_PERCENT; In general you can ignore anything with a p-Value > 0.05 and the things with very low p-Values (< 2e-16) have the most certain effects.  'Estimate' is the effect size; -2.4 means that for each step from the age group 18-29 to the age group 65+ reduces the percentage of mental distress by 2.4% after accounting for all of the other variables included in the model.  

The R2 value is a measure of goodness of fit.  0.166 means that about 16.6% of the variance in MENTAL_DISTRESS_PERCENT is explained by this model.  In other words, about 83.4% of a person's mental distress is caused by other things.

A powerful tool for removing variables that are caused by other variables is LASSO regression.  For example if the effect of age was due to being married or not, LASSO would remove the age variable.
```{r Covariates.lasso}
x <- mar20Dat.curDF[ , strsplit(as.character(fullEq)[3], ' \\+ ')[[1]]]
y <- mar20Dat.curDF$MENTAL_DISTRESS_PERCENT 
lasso.cv <- cv.glmnet(x=as.matrix(x), y=y)

message('These parameters survive LASSO regression to predict MHEALTH')
extract.coef(lasso.cv)

predVal <- predict(lasso.cv, newx=as.matrix(x))
r2 <- cor(predVal, y)^2
message(paste('R2 =', round(r2,3)))
```
In these LASSO results you can see

-2.358040667	NA	F_AGECAT
which essentially repeate the previous results, where each step from the age group 18-29 to the age group 65+ reduces the percentage of mental distress by ~2.4% after accounting for all of the other variables included in the model. 

2.841406376	NA	F_IDEO
Shows that after controlling for the other variables, each step from "1: Very conservative" to "5: Very liberal" increases mental distress by about 2.8%

Note that this analysis is limited in that it's using linear models, which might not be appropriate.  For example F_HOW_MUCH_MARRIED considers 'Widowed' as one point less married than 'Separated' which is 3 points less married than 'Married'.  In other words: there is plenty of room for improvement.

## Step 6: Replicate on another data set

For follow-up, use W83 from Feb 21.  That also asks for political party and signs of mental health stress.

```{r  LoadFollowup.W83}
feb21Dat <- haven::read_sav(file=file.path('W83_Feb21/', 'ATP W83.sav'))
```

```{r LoadRelevantVars.W83}
#Note: Race will need to be one-hot
relCov <- c("F_AGECAT", "F_MARITAL", "F_EDUCCAT2", "F_INC_SDT1",  "F_IDEO", "F_ATTEND")  #"MHEALTH", "F_RACETHN" , "F_SEX", "COVID_MENTAL_W64"

feb21DF <- as.data.frame(feb21Dat[, c(relCov, "MHEALTH_W83", "F_RACECMB", "F_GENDER")])
feb21DF[feb21DF==99] <- as.numeric(NA)
feb21DF <- feb21DF[!apply(feb21DF, 1, function(x){any(is.na(x))}),]

#One-hot race
F_RACETHN_Name <- names(attr(feb21Dat$F_RACECMB, "labels"))[match(feb21DF$F_RACECMB, attr(feb21Dat$F_RACECMB, "labels"))]
table(F_RACETHN_Name)
oHot_Race <- fastDummies::dummy_cols(F_RACETHN_Name, remove_selected_columns=TRUE)
colnames(oHot_Race) <- make.names(sub('\\.data_', '', colnames(oHot_Race)))

#One-hot sex
F_SEX_Name <- names(attr(feb21Dat$F_GENDER, "labels"))[match(feb21DF$F_GENDER, attr(feb21Dat$F_GENDER, "labels"))]
table(F_SEX_Name)
oHot_Sex <- fastDummies::dummy_cols(F_SEX_Name, remove_selected_columns=TRUE)
colnames(oHot_Sex) <- make.names(sub('\\.data_', '', colnames(oHot_Sex)))

oHots <- cbind(oHot_Race, oHot_Sex)

#Fix religion and marriage so higher is better
feb21DF <- cbind(feb21DF, oHots)
feb21DF$F_HOW_MUCH_RELIGION <- 7-feb21DF$F_ATTEND

feb21DF$F_HOW_MUCH_MARRIED <- 7-feb21DF$F_MARITAL

#Fix MHEALTH so it varies by percentage of mental distress
feb21DF$MENTAL_DISTRESS_PERCENT <- round(100*(feb21DF$MHEALTH_W83 - min(feb21DF$MHEALTH_W83))/(max(feb21DF$MHEALTH_W83)- min(feb21DF$MHEALTH_W83)))
```

```{r Covariates.regression.W83}
eq1 <- sub('F_MARITAL', 'F_HOW_MUCH_MARRIED', sub('F_ATTEND', 'F_HOW_MUCH_RELIGION', paste(relCov, collapse=' + ')))
eq2 <- paste(colnames(oHots)[!colnames(oHots) %in% c('Other', 'Male')], collapse=' + ')
fullEq.str <- paste0('MENTAL_DISTRESS_PERCENT ~ ', eq1, " + ", eq2)

message('Prediction of MENTAL_DISTRESS_PERCENT aggregate variable')
message(fullEq.str)
fullEq <- as.formula(fullEq.str)

fullReg <- glm(formula = fullEq, data = feb21DF)
summary(fullReg)

regPred <- predict(fullReg)
r2 <- cor(regPred, feb21DF$MENTAL_DISTRESS_PERCENT)^2
message(paste('R2 =', round(r2,3)))
```

Note that this analysis has an R2 value ~5% lower than the previous model.  This is probably due to the missing question about diagnosed mental illness.  If you remove that variable from the previous model, that R2 value drops to around 0.117.


```{r Covariates.lasso.W83}
x <- feb21DF[ , strsplit(as.character(fullEq)[3], ' \\+ ')[[1]]]
y <- feb21DF$MENTAL_DISTRESS_PERCENT 
lasso.cv <- cv.glmnet(x=as.matrix(x), y=y)

message('These parameters survive LASSO regression to predict MHEALTH')
extract.coef(lasso.cv)

predVal <- predict(lasso.cv, newx=as.matrix(x))
r2 <- cor(predVal, y)^2
message(paste('R2 =', round(r2,3)))
```
This LASSO regression of the February ;21 data shows very similar results to those presented in the March '20 data.

One important difference: 
1.6470000	NA	F_IDEO

Shows that after controlling for the other variables, each step from "1: Very conservative" to "5: Very liberal" increases mental distress by about 1.6% compared to the 2.8% reported previously.  Perhaps this reflects a Democratic victory in the 2020 presidential election.

```{r IdeologyByParty}
F_PARTY_FINAL_name <- names(attr(feb21Dat$F_PARTY_FINAL, "labels"))[match(feb21Dat$F_PARTY_FINAL, attr(feb21Dat$F_PARTY_FINAL, "labels"))]
F_PARTY_IDEO_name <- names(attr(feb21Dat$F_IDEO, "labels"))[match(feb21Dat$F_IDEO, attr(feb21Dat$F_IDEO, "labels"))]
F_PARTY_IDEO_name <- factor(F_PARTY_IDEO_name , levels=names(attr(feb21Dat$F_IDEO, "labels")))

table(F_PARTY_IDEO_name, F_PARTY_FINAL_name)
```

## Step 7: Conclusion

Taken as a whole, this analysis shows that people who identify as 'Liberal' or 'Very Liberal' much more mentally ill / distressed than they rest of the population.  Additionally, after controlling for other variables, a liberal ideology remains associated with increased mental distress.



